Combining global tree cover loss data with historical national forest-cover

### 2018-01-16 manuscript/manuscript.Rmd
Bookdown

---
title: "Combining global tree cover loss data with historical national forest-cover maps to look at six decades of deforestation and forest fragmentation in Madagascar"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::word_document2: null
  bookdown::pdf_document2:
    #includes:
    #  before_body: doc_prefix.tex
    #  in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    toc: no
bibliography: bib/biblio.bib
fontsize: 12pt
link-citations: yes
csl: bib/journal-of-applied-ecology.csl
---

```{r load_data}
load("deforestmap.rda")
```

#### Ghislain Vieilledent<sup>*,1,2</sup>, Clovis Grinand<sup>3</sup>, Fety A. Rakotomalala<sup>3</sup>, Rija Ranaivosoa<sup>4</sup>, Jean-Roger Rakotoarijaona<sup>4</sup>, Thomas F. Allnutt<sup>5,6</sup>, and Frédéric Achard<sup>1</sup>

@@

## Highlights

@@

## Summary

@@

## Introduction

@@

## Materials and Methods

### Creation of new forest cover maps of Madagascar from 1953 to 2014

@@

### Computing forest cover areas and deforestation rates

@@

\begin{equation}
  \theta = 100 \times [1-(1-(F_{t_2}-F_{t_1})/F_{t_1})^{(1/(t_2-t_1))}]
  (\#eq:theta)
\end{equation}

In Eq. \ref{eq:theta}, $\theta$ is the annual deforestation rate (in
%/yr), $F_{t_2}$ and $F_{t_1}$ are the forest cover free of clouds at
both dates $t_2$ and $t_1$, and $t_2-t_1$ is the time-interval (in
years) between the two dates.

@@

### Comparing our forest cover and deforestation rate estimates with previous studies

We compared our estimates of forest cover and deforestation rates with
estimates from the three existing studies at the national scale for
Madagascar: (i) [@Harper2007], (ii) [@MEFT2009] and (iii) [@ONE2015].
@Harper2007 provides forest cover and deforestation estimates for the
periods c. 1953-c. 1973-1990-2000. @MEFT2009 provides estimates for the
periods 1990-2000-2005 and @ONE2015 provides estimates for the periods
2005-2010-2013. To compare our forest cover and deforestation estimates
over the same time periods, we consider an additional time-period in our
study (2010-2013) by creating an extra forest cover map for the year
2013. We computed the Pearson's correlation coefficient and the root
mean square error (RMSE) between our forest cover estimates and forest
cover estimates from previous studies for all the dates and forest types
(including also the total forest cover estimates). For previous studies,
the computation of annual deforestation rates (in %/yr) is not always
detailed and might slightly differ from one study to another [see @Puyravaud2003]. @Harper2007 also provide total deforested areas for the
two periods 1973-1990 and 1990-2000. We converted these values into
annual deforested area estimates. When annual deforested areas were not
reported (for 1953-1973 in @Harper2007 and in @MEFT2009 and @ONE2015),
we computed them from the forest cover estimates in each study. These
estimates cannot be corrected from the potential bias due to the
presence of residual clouds. forest cover and deforestation rates were
then compared between all studies for the whole of Madagascar and the
four ecoregions. The same ecoregion boundaries as in our study were used
in @ONE2015 but this was not the case for @Harper2007 and @MEFT2009,
which can explain a part of the differences between the estimates.

### Fragmentation

@@

## Results

### Forest cover change and deforestation rates

@@

### Comparison with previous forest cover change studies in Madagascar

@@

### Evolution of forest fragmentation with time

@@

## Discussion

### Advantages of combining recent global annual tree cover loss data with historical national forest cover maps

@@

### Natural forest cover change in Madagascar from 1953 to 2014

@@

### Deforestation trend and impacts on conservation and development policies

@@

## Author's contribution

@@

## Acknowledgements

@@

## Data accessibility

@@

## Tables

```{r forest_cover, results="asis"}
defor.df <- read.table(file="tabs/defor.txt", sep="\t", header=TRUE)
perc.1 <- round(100*(defor.df$area[defor.df$Year==1973]-defor.df$area[defor.df$Year==2014])/defor.df$area[defor.df$Year==1973])
perc.2 <- round(100*(defor.df$area[defor.df$Year==1953]-defor.df$area[defor.df$Year==2014])/defor.df$area[defor.df$Year==1953])
d.1 <- defor.df$theta[defor.df$Year==1990]
d.2 <- defor.df$theta[defor.df$Year==2010]
tab.1 <- defor.df
tab.1[,c(2:4)] <- round(tab.1[,c(2:4)]/1000)
tab.1$Year <- as.character(tab.1$Year)
names(tab.1) <- c("Year", "Forest (Kha)", "Unmap (Kha)", "Annual defor. (Kha/yr)", "Rate (%/yr)")
kable(tab.1, format.args = list(big.mark=","), digits=1)
```

Table 1: **Evolution of natural forest cover and deforestation rates from 1953 to 2014 in Madagascar.** Areas are provided in thousands of hectares (Kha). Forest map for the year 1973 has 3.3 Mha of unclassified areas due to the presence of clouds on satellite images. As a consequence, partial deforestation rates for the periods 1953-1973 and 1973-1990 are computed based on the available forest extent. The last two columns indicate the annual deforested areas and annual deforestation rates on the previous time-period (e.g. 1953-1973 for year 1973, 1973-1990 for year 1990, etc.).

```{r comp_forest, results="asis"}
fcc.comp <- read.table(file="tabs/fcc_comp.txt", sep=" ", header=TRUE)
tab.2 <- fcc.comp
tab.2 <- tab.2[,c(2,1,3:10)]
tab.2[,c(3:10)] <- round(tab.2[,c(3:10)]/1000)
tab.2$ForestType <- c("Total",rep("",3),"Moist",rep(" ",3),"Dry",rep(" ",3),"Spiny",rep(" ",3),"Mangroves",rep(" ",3))
col.names <- c("Forest type","Source","1953","1973","1990","2000","2005","2010","2013","2014")
kable(tab.2,col.names=col.names,align=c("l","l",rep("r",8)), format.args = list(big.mark=","))
```

Table 2: **Comparing Madagascar forest cover estimates with previous studies on the period 1953-2014.** We compared our estimates of forest cover with the estimates from three previous studies [@Harper2007; @MEFT2009; @ONE2015]. Areas are provided in thousands of hectares (Kha). We obtained a Pearson's correlation coefficient of 0.99 between our forest cover estimates and forest cover estimates from previous studies. The increase in mangrove and spiny forest covers from _c._ 1953 to _c._ 1973 in @Harper2007 and our study is most probably due to differences in forest definition and mapping methods between the 1953 aerial-photography derived map and the 1973 Landsat image derived map.

```{r comp_defor, results="asis"}
tab.3 <- read.table(file="tabs/defor_comp.txt", sep=" ", header=TRUE)
tab.3$ForestType <- c("Total",rep("",3),"Moist",rep(" ",3),"Dry",rep(" ",3),"Spiny",rep(" ",3),"Mangroves",rep(" ",3))
col.names <- c("Forest type","Source","1953-1973","1973-1990","1990-2000","2000-2005","2005-2010","2010-2013")
kable(tab.3,col.names=col.names,align=c("l","l",rep("r",6)))
```

Table 3: **Comparing Madagascar annual deforestation rates with previous studies on the period 1953-2014.** Annual deforested areas (in thousands of hectares per year, Kha/yr) and annual deforestation rates (second number in parenthesis, in %/yr) are provided. For deforestation rates in %/yr, exact same numbers as in scientific articles and reports from previous studies [@Harper2007; @MEFT2009; @ONE2015] have been reported. The way annual deforestation rates in %/yr have been computed in these previous studies can slightly differ from one study to another, but estimates always correct for the potential presences of clouds on satellite images and unclassified areas on forest maps. Annual deforested areas in Kha/yr have been recomputed from forest cover estimates in Tab. \ref{tab:comp_forest} (except for @Harper2007 for the periods 1973-1990 and 1990-2000 for which annual deforested areas in Kha/yr were derived from numbers reported in the original publication, see methods) and do not correct for the potential presence of clouds.

```{r frag, results="asis"}
frag.df <- read.table(file="outputs/fordens.txt", sep="\t", header=TRUE)
tab.4 <- frag.df
p.1 <- round(frag.df$interior[frag.df$Year==1973])
p.2 <- round(frag.df$interior[frag.df$Year==2014])
p.3 <- round(frag.df$patch[frag.df$Year==2014]+frag.df$transitional[frag.df$Year==2014])
tab.4 <- tab.4[,-c(8)]
tab.4$Year <- as.character(tab.4$Year)
tab.4$forest <- round(tab.4$forest/1000)
names(tab.4) <- c("Year","Forest (ha)", "0-20", "21-40", "41-60", "61-80", "81-100")
kable(tab.4, digits=0, format.args=list(big.mark=","))
```

Table 4: **Evolution of the forest fragmentation from 1953 to 2014 in Madagascar.** Five forest fragmentation classes, based on the percentage of forest in the neighborhood, are defined: 0-20% (highly fragmented), 21-40%, 41-60%, 61-80% and 81-100% (lowly fragmented). The percentage of forest falling in each forest fragmentation class is reported for each year. Forest areas are provided in thousands of hectares (Kha).

##  Figures

@@

## References

>>>>

---
title: Combining global tree cover loss data with historical national forest-cover
  maps to look at six decades of deforestation and forest fragmentation in Madagascar

output:
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: yes
    toc: no

link-citations: yes
bibliography: bib/biblio.bib
csl: bib/journal-of-applied-ecology.csl
---

```{r options, echo=FALSE, results="hide"}
library(bookdown)
library(knitr)
library(kableExtra)
opts_chunk$set(echo=FALSE, cache=FALSE,
               #results="hide", 
               warning=FALSE,
               message=FALSE, highlight=TRUE,
               fig.show="hide", size="small",
               tidy=FALSE)
options(knitr.kable.NA="-")
```

**Ghislain Vieilledent$^{1,2,\star}$, Clovis Grinand$^{3}$, Fety A. Rakotomalala$^{3}$, Rija Ranaivosoa$^{4}$, Jean-Roger Rakotoarijaona$^{4}$, Thomas F. Allnutt$^{5,6}$, and Frédéric Achard$^{1}$**

@@

# Highlights

@@

# Summary

@@

# Introduction

@@

# Materials and Methods

## Creation of new forest cover maps of Madagascar from 1953 to 2014

@@

## Computing forest cover areas and deforestation rates

@@

\begin{equation}
  (\#eq:theta)
  \theta = 100 \times [1-(1-(F_{t_2}-F_{t_1})/F_{t_1})^{(1/(t_2-t_1))}]
\end{equation}

In Eq. \@ref(eq:theta), $\theta$ is the annual deforestation rate (in
%/yr), $F_{t_2}$ and $F_{t_1}$ are the forest cover free of clouds at
both dates $t_2$ and $t_1$, and $t_2-t_1$ is the time-interval (in
years) between the two dates.

@@

## Comparing our forest cover and deforestation rate estimates with previous studies

We compared our estimates of forest cover and deforestation rates with
estimates from the three existing studies at the national scale for
Madagascar: (i) @Harper2007, (ii) @MEFT2009 and (iii) @ONE2015.
@Harper2007 provides forest cover and deforestation estimates for the
periods c. 1953-c. 1973-1990-2000. @MEFT2009 provides estimates for the
periods 1990-2000-2005 and @ONE2015 provides estimates for the periods
2005-2010-2013. To compare our forest cover and deforestation estimates
over the same time periods, we consider an additional time-period in our
study (2010-2013) by creating an extra forest cover map for the year
2013. We computed the Pearson's correlation coefficient and the root
mean square error (RMSE) between our forest cover estimates and forest
cover estimates from previous studies for all the dates and forest types
(including also the total forest cover estimates). For previous studies,
the computation of annual deforestation rates (in %/yr) is not always
detailed and might slightly differ from one study to another [see @Puyravaud2003]. @Harper2007 also provide total deforested areas for the
two periods 1973-1990 and 1990-2000. We converted these values into
annual deforested area estimates. When annual deforested areas were not
reported (for 1953-1973 in @Harper2007 and in @MEFT2009 and @ONE2015),
we computed them from the forest cover estimates in each study. These
estimates cannot be corrected from the potential bias due to the
presence of residual clouds. Forest cover and deforestation rates were
then compared between all studies for the whole of Madagascar and the
four ecoregions. The same ecoregion boundaries as in our study were used
in @ONE2015 but this was not the case for @Harper2007 and @MEFT2009,
which can explain a part of the differences between the estimates.

## Fragmentation

@@

# Results

## Forest cover change and deforestation rates

@@

## Comparison with previous forest cover change studies in Madagascar

@@

## Evolution of forest fragmentation with time

@@

# Discussion

## Advantages of combining recent global annual tree cover loss data with historical national forest cover maps

@@

## Natural forest cover change in Madagascar from 1953 to 2014

@@

## Deforestation trend and impacts on conservation and development policies

@@

# Author's contribution

@@

# Acknowledgements

@@

# Data accessibility

@@

# Tables

```{r forest_cover}
defor.df <- read.table(file="tabs/defor.txt", sep="\t", header=TRUE)
perc.1 <- round(100*(defor.df$area[defor.df$Year==1973]-defor.df$area[defor.df$Year==2014])/defor.df$area[defor.df$Year==1973])
perc.2 <- round(100*(defor.df$area[defor.df$Year==1953]-defor.df$area[defor.df$Year==2014])/defor.df$area[defor.df$Year==1953])
d.1 <- defor.df$theta[defor.df$Year==1990]
d.2 <- defor.df$theta[defor.df$Year==2010]
tab.1 <- defor.df
tab.1[,c(2:4)] <- round(tab.1[,c(2:4)]/1000)
tab.1$Year <- as.character(tab.1$Year)
names(tab.1) <- c("Year", "Forest (Kha)", "Unmap (Kha)", "Annual defor. (Kha/yr)", "Rate (%/yr)")
# Caption
#cap <- "**Evolution of natural forest cover and deforestation rates from 1953 to 2014 in Madagascar.** Areas are provided in thousands of hectares (Kha). Forest map for the year 1973 has 3.3 Mha of unclassified areas due to the presence of clouds on satellite images. As a consequence, partial deforestation rates for the periods 1953-1973 and 1973-1990 are computed based on the available forest extent. The last two columns indicate the annual deforested areas and annual deforestation rates on the previous time-period (e.g. 1953-1973 for year 1973, 1973-1990 for year 1990, etc.)."
# Kable
knitr::kable(tab.1, format.args=list(big.mark=","), digits=1, booktabs=TRUE, caption="Test")
```

```{r comp_forest}
fcc.comp <- read.table(file="tabs/fcc_comp.txt", sep=" ", header=TRUE)
tab.2 <- fcc.comp
tab.2 <- tab.2[,c(2,1,3:10)]
tab.2[,c(3:10)] <- round(tab.2[,c(3:10)]/1000)
tab.2$ForestType <- c("Total",rep("",3),"Moist",rep(" ",3),"Dry",rep(" ",3),"Spiny",rep(" ",3),"Mangroves",rep(" ",3))
col.names <- c("Forest type","Source","1953","1973","1990","2000","2005","2010","2013","2014")
# Caption
#cap <- "**Comparing Madagascar forest cover estimates with previous studies on the period 1953-2014.** We compared our estimates of forest cover with the estimates from three previous studies [@Harper2007; @MEFT2009; @ONE2015]. Areas are provided in thousands of hectares (Kha). We obtained a Pearson's correlation coefficient of 0.99 between our forest cover estimates and forest cover estimates from previous studies. The increase in mangrove and spiny forest covers from _c._ 1953 to _c._ 1973 in @Harper2007 and our study is most probably due to differences in forest definition and mapping methods between the 1953 aerial-photography derived map and the 1973 Landsat image derived map."
# Kable
knitr::kable(tab.2, col.names=col.names,align=c("l","l",rep("r",8)), format.args=list(big.mark=","), booktabs=TRUE, caption="Test")
```

```{r comp_defor}
tab.3 <- read.table(file="tabs/defor_comp.txt", sep=" ", header=TRUE)
tab.3$ForestType <- c("Total",rep("",3),"Moist",rep(" ",3),"Dry",rep(" ",3),"Spiny",rep(" ",3),"Mangroves",rep(" ",3))
col.names <- c("Forest type","Source","1953-1973","1973-1990","1990-2000","2000-2005","2005-2010","2010-2013")
# Caption
# cap <- "**Comparing Madagascar annual deforestation rates with previous studies on the period 1953-2014.** Annual deforested areas (in thousands of hectares per year, Kha/yr) and annual deforestation rates (second number in parenthesis, in %/yr) are provided. For deforestation rates in %/yr, exact same numbers as in scientific articles and reports from previous studies [@Harper2007; @MEFT2009; @ONE2015] have been reported. The way annual deforestation rates in %/yr have been computed in these previous studies can slightly differ from one study to another, but estimates always correct for the potential presences of clouds on satellite images and unclassified areas on forest maps. Annual deforested areas in Kha/yr have been recomputed from forest cover estimates in Tab. \\@ref(tab:comp_forest) (except for @Harper2007 for the periods 1973-1990 and 1990-2000 for which annual deforested areas in Kha/yr were derived from numbers reported in the original publication, see methods) and do not correct for the potential presence of clouds."
# Kable
knitr::kable(tab.3, col.names=col.names, align=c("l","l",rep("r",6)), booktabs=TRUE, caption="Test")
```

```{r frag}
# Import and format
frag.df <- read.table(file="tabs/fordens.txt", sep="\t", header=TRUE)
tab.4 <- frag.df
tab.4$Year <- as.character(tab.4$Year)
tab.4$forest <- round(tab.4$forest/1000)
names(tab.4) <- c("Year","Forest (Kha)", "0-20", "21-40", "41-60", "61-80", "81-100")
# Caption
#cap <- "**Evolution of the forest fragmentation from 1953 to 2014 in Madagascar.** Five forest fragmentation classes, based on the percentage of forest in the neighborhood, are defined: 0-20% (highly fragmented), 21-40%, 41-60%, 61-80% and 81-100% (lowly fragmented). The percentage of forest falling in each forest fragmentation class is reported for each year. Forest areas are provided in thousands of hectares (Kha)."
# Kable
knitr::kable(tab.4, digits=0, format.args=list(big.mark=","), booktabs=TRUE, caption="Test")
```

#  Figures

@@

# References
